---
title: "AVR"
output:
html_document:
df_print: paged
toc: yes
toc_float:
collapsed: yes
smooth_scroll: no
bibliography: bibliography.bib
nocite: '@*'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(tidyverse, ggpubr)
```

# Introduction

Data was gathered using questionnaires and mark reports. Mark reports were provided
by schools and consist of grades and absences. Grades are evaluated in three periods
and marked in a 20-point grading scale where 0 is the lowest possible mark and
20 the highest. The former was used to gather the data for the several demographic,
social/emotional and school related variables. The final questionnaire containing
37 questions in a single A4 sheet was answered in class by 788 students. Later
111 answers were discarded due to lack of identification details. At the end the
data was merged into two samples that related to Mathematics (395) and
Portuguese (649) classes.

```{r}
source("read_data.R")
master_data <- read_data()
```

# Question 1
Getting into the weeds of student’s health among two schools might reveal their well-being.

When analyzing the dataset it confused us that there are two separate columns for health.
We were convinced that this should be gathered only once during the research. We have
decided to check how sitting in the particular subject influences the responses.

```{r}

health_eq_acc_points <- master_data %>%
        mutate(health_eq = if_else(health_m == health_p, 1, 0)) %>%
        group_by(health_m, health_p) %>%
        summarise(cnt = n())

```

```{r}
ggplot(master_data %>%
               inner_join(health_eq_acc_points),
       aes(x = health_m, y = health_p, size = cnt)) +
        scale_size(name   = "",
                   breaks = c(144, 61, 1),
                   labels = c('144', '[42, 80]', '[1, 2]')) +
        geom_point() +
        labs(y = 'Health reported in the math classes',
             x = 'Health reported in the portugese classes',
             title = 'Relationship between reported health at Math and Portguese classess')
```
# Question 2
What kind of support students with each subject? 

```{r barplots}

sum(master_data$paid_m != master_data$paid_p)
# subject support from family

master_data %>% select(famsup_m,famsup_p) %>% mutate(n=1) %>% pivot_longer(!n) %>% 
  ggplot(aes(x = value, fill = name)) +
  geom_bar(stat = "count", position = "dodge") +
  scale_fill_discrete(name = "Subject", labels = c("Maths", "Portuguese"))+
  ggtitle("Does student obtain family support?")+
  xlab(NULL)

# paid support
master_data %>% select(paid_m,paid_p) %>% mutate(n=1) %>% pivot_longer(!n) %>% 
  ggplot(aes(x = value, fill = name)) +
  geom_bar(stat = "count", position = "dodge")+
  scale_fill_discrete(name = "Subject", labels = c("Maths", "Portuguese"))+
  ggtitle("Does student obtain paid support?")+
  xlab(NULL)

# school support
master_data %>% select(schoolsup_m,schoolsup_p) %>% mutate(n=1) %>% pivot_longer(!n) %>% 
  ggplot(aes(x = value, fill = name)) +
  geom_bar(stat = "count", position = "dodge")+
  scale_fill_discrete(name = "Subject", labels = c("Maths", "Portuguese"))+
  ggtitle("Does student obtain school support?")+
  xlab(NULL)

```


```{r}
# better varsion of barplot to answer the quesition

# create additional variables

master_data %>% group_by(famsup_m,schoolsup_m,paid_m) %>% 
  summarise(n = n(), grade = mean(G3_m)) %>% 
  arrange(-grade)%>% pivot_longer(c(famsup_m,schoolsup_m,paid_m)) %>% 
  mutate(bg_color = ifelse(value == "yes", "green","red")) ->tmp_1

tiles_df <- data.frame(
  x = rep(seq(2,16,2), 3),
  y = rep(c(1,2,3), each = 8)
) %>% arrange(x) %>% mutate(z = tmp_1$value)
ggplot() +
  geom_tile(data = tiles_df, aes(x, y,fill = z), colour = "grey50", alpha = 0.3)+
  scale_fill_manual(values=c("#a3d487", "#f21a02"))+
  scale_y_continuous(labels = c("a","b","c"), breaks = c(1,2,3))
```


Most students reported the same health condition in Math and Portugese classes.

# References
