---
title: "Advanced Visualization in R"
subtitle: Using advanced visualization techniques to explore relationships between students' characteristics & performance.
author: "Wojciech Maślakiewicz & Mateusz Baryła"
output:
  html_document:
      code_folding: hide
      df_print: paged
      toc: yes
      toc_float: yes
      collapsed: yes
      smooth_scroll: no
bibliography: bibliography.bib
nocite: '@*'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE, warning = FALSE)
pacman::p_load(tidyverse, readxl, ggpubr, riverplot,
               gridExtra, grid, tidyverse, treemapify,
               ggradar, ggstance, ggthemes,ggcharts)
color_palette <- tibble(colors = c("#F5F5F5", "#F05454", "#30475E", "#121212"),
                        color_names = c("grey", "red", "blue", "black"))
```

# Introduction

Data was gathered using questionnaires and mark reports. Mark reports were provided
by schools and consist of grades and absences. Grades are evaluated in three periods
and marked in a 20-point grading scale where 0 is the lowest possible mark and
20 the highest. The former was used to gather the data for the several demographic,
social/emotional and school related variables. The final questionnaire containing
37 questions in a single A4 sheet was answered in class by 788 students. Later
111 answers were discarded due to lack of identification details. At the end the
data was merged into two samples that related to Mathematics (395) and
Portuguese (649) classes [@Cortez2008].

The merged dataset consist of 13 variables that uniquely identifies
each row and 20 that describe their performance at one of the subjects.
All the features can be split into categorical and numerical ones, but we
are rather interested in analyzing the variables rather than the right survey distribution [@Dua:2019].

```{r, echo = FALSE}
source("read_data.R")
source("plot_functions.R")
master_data <- read_data()
dictionary <- read_excel("dictionary.xlsx")
```

# Question 1

## What kind of support students get for each subject?

```{r }
supp_plot_m <- make_barplot_with_bg_tiles(master_data,famsup_m,schoolsup_m,paid_m,G3_m,
                                          title = '',
                                          bar_color = "#30475E")

supp_plot_p <- make_barplot_with_bg_tiles(master_data,famsup_p,schoolsup_p,paid_p,G3_p,
                                          title = '',
                                          bar_color = "#30475E")

supp_plot_p <- supp_plot_p + theme(axis.line.y.right = element_blank(),
                                   axis.text.y.right = element_blank(),
                                   axis.ticks.y.right = element_blank(),
                                   axis.title.y.right = element_blank(),
                                   axis.title.y.left = element_text(size=18),
                                   axis.text.y.left = element_text(size=19))

supp_plot_m <- supp_plot_m + theme(axis.line.y.left = element_blank(),
                                   axis.text.y.left = element_blank(),
                                   axis.ticks.y.left = element_blank(),
                                   axis.title.y.left = element_blank(),
                                   axis.title.y.right = element_text(size=18),
                                   axis.text.y.right = element_text(size=19))
comb <- ggarrange(supp_plot_p,supp_plot_m,
                  align = "h",
                  labels = c("Portuguese","Maths"),
                  font.label = list(size = 12, face = 'plain'),
                  hjust = c(-2,-2.5),
                  vjust = 3,
                  common.legend = T ,
                  legend = "bottom")

annotate_figure(comb , top = text_grob("Does obtaining support improve performance?",
                                       size = 23))

```

## Insight

What we observe is that in Portuguese classes most students get only family 
support or no support at all. Interestingly, at the same time those students 
have the highest average grade. This would suggest that support is given to 
students who tend to perform poorly in class rather than students perform well
BECAUSE they got support.

In Maths most students get family and paid support or no support. The highest 
average is among those getting paid support and both paid and family support.

It's worth mentioning that the small group of students getting school support and
paid support at the same time has a close-to-the-lowest average from both subjects. 


# Question 2

## Do students improve their grades during the year?

```{r}
# data
grade_cuts <-  c(-1,7,13,21)
tmp1 <- master_data %>% 
  select(matches("G._m")) %>% 
  drop_na()%>% 
  mutate(across(everything(),
                ~cut(.x,breaks = grade_cuts,
                    labels = c("low","medium","high")))) %>% 
  group_by(G1_m,G2_m,G3_m) %>%
  summarise(n = n()) %>% 
  mutate(G1_m = paste0("First ",G1_m),
         G2_m = paste0("Second ",G2_m),
         G3_m = paste0("Final ",G3_m)) %>% 
  arrange(G1_m,G2_m)



# final plot
nodes <- data.frame(ID = unique(c(tmp1$G1_m,tmp1$G2_m,tmp1$G3_m) %>% sort()),
                    x = c(rep(3,3),rep(1,3),rep(2,3)),
                    y = rep(c(3,1,2),3))

tmp2 <- master_data %>% 
  select(matches("G._m")) %>% 
  drop_na()%>% 
  mutate(across(everything(),
                ~cut(.x,breaks =grade_cuts,
                    labels = c("low","medium","high")))) %>% 
  group_by(G1_m,G2_m) %>% 
  summarise(n = n()) %>% 
  mutate(G1_m = paste0("First ",G1_m),
         G2_m = paste0("Second ",G2_m)) %>% 
  rename(N1 = G1_m, N2 = G2_m, Value = n)

tmp3 <- master_data %>% 
  select(matches("G._m")) %>% 
  drop_na()%>% 
  mutate(across(everything(),
                ~cut(.x,breaks = grade_cuts,
                    labels = c("low","medium","high")))) %>% 
  group_by(G2_m,G3_m) %>% 
  summarise(n = n()) %>% 
  mutate(G2_m = paste0("Second ",G2_m),
         G3_m = paste0("Final ",G3_m)) %>% 
  rename(N1 = G2_m, N2 = G3_m, Value = n)

tmp4 <- rbind(tmp2,tmp3)

edges <- data.frame(ID = 1:nrow(tmp4),
                    N1 = tmp4$N1,
                    N2 = tmp4$N2,
                    Value = tmp4$Value)
cols <- c(low="#F05454",
medium="#30475E",
high ="#121212")
style <- sapply(nodes$ID, function(id)
list(col=cols[ gsub("(First|Second|Final) ", "", id) ]), simplify=FALSE)



r <- makeRiver(nodes,edges,styles =style) 
d <- list(srt = 0, textcex=1, textcol = "white") 

plot(r, plot_area=1, nodewidth=10, default_style=d)
title("Flows of students between performance groups")
```

## Insight 

We observe that the performace of students is rather persistant across the year.
Very little students tend to either improve or worsen their performance.

The biggest observable flows between performance groups happens between first and 
second period especially between low and medium group. Nonetheless, those flows 
are modest and seem not to change the overall distribution of performance where
the most students are the medium ones.

# Question 3

## What is the distribution of grades across subjects and periods in the current school year?

```{r}
grade_per_subject_long <-
  master_data %>%
  mutate(row_num = row_number()) %>%
  select(row_num, starts_with("G", ignore.case = FALSE)) %>%
  pivot_longer(-row_num) %>%
  separate(name, c("grade", "subject"))
subject_pretty_names <- tibble(short_name = c('m', 'p'),
                               pretty_name = c('Mathematics','Portugese'))
periods_colors <- tibble(period = 1:3,
                         color_name = color_palette$colors[2:4])

data_to_boxplot <- grade_per_subject_long %>%
  inner_join(subject_pretty_names, by = c("subject" = "short_name")) %>% separate(grade, 1, into = c("to_remove", "period")) %>%
  mutate(period = as.numeric(period)) %>%
  inner_join(periods_colors)
```

```{r}
grades_boxplot <- ggplot(data = data_to_boxplot,
       aes(y = value, x = factor(pretty_name),
           fill = factor(period),
           colour=factor(period))) +
  geom_boxplot(alpha = 0.6, show.legend = F) +
  geom_point(position = position_jitterdodge()) +
  ggtitle("What is the distribution of grades across \n subjects and periods in the current school year?") +
  labs(x = "Subject", y = "Grade", colour = "Period") +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_color_manual(values = periods_colors$color_name) +
  scale_fill_manual(values = periods_colors$color_name) +
  guides(fill=FALSE)
```


```{r}
grades_boxplot
```

## Insight 

Grades for Portugese are consistently higher than in Mathematics.
We do not observe neither a positive nor negative trend relating
to the subsequent periods results. However, there are significantly more students
that failed Mathematics.

# Question 4

## How the amount of spare time and the way of spending it influence the number of absences?


```{r}
grade_per_subject_long <-
  master_data %>%
  mutate(row_num = row_number()) %>%
  select(row_num, starts_with("G", ignore.case = FALSE)) %>%
  pivot_longer(-row_num) %>%
  separate(name, c("grade", "subject"))

absences.agg <- master_data %>%
  group_by(freetime_m, goout_m) %>%
  summarise(across(absences_m, ~n()))

low.high.map <- tibble(ugly_level = c(1, 2, 3, 4, 5),
                         pretty_level = c('very low',
                                               'low',
                                               'medium',
                                               'high',
                                               'very high'))

absences.agg.to.plot.m <- absences.agg %>%
  ungroup() %>%
  inner_join(low.high.map,
             by = c("freetime_m" = "ugly_level")) %>%
  rename( freetime_pretty = pretty_level) %>%
  inner_join(low.high.map,
             by = c("goout_m" = "ugly_level")) %>%
  rename( goout_pretty = pretty_level) %>% select(freetime_pretty, goout_pretty, absences_m)

absences.agg <- master_data %>%
  group_by(freetime_p, goout_p) %>%
  summarise(across(.cols = absences_p, .fns = ~sum(.)))

low.high.map <- tibble(ugly_level = c(1, 2, 3, 4, 5),
                       pretty_level = c('very low',
                                        'low',
                                        'medium',
                                        'high',
                                        'very high'))
absences.agg.to.plot.p <- absences.agg %>%
  ungroup() %>%
  inner_join(low.high.map,
             by = c("freetime_p" = "ugly_level")) %>%
  rename(freetime_pretty = pretty_level) %>%
  inner_join(low.high.map,
             by = c("goout_p" = "ugly_level")) %>%
  mutate(goout_p = factor(goout_p)) %>%
  rename( goout_pretty = pretty_level) %>%
  select(freetime_pretty, goout_pretty, absences_p)

absences.agg.to.plot.p$freetime_pretty <- factor(absences.agg.to.plot.p$freetime_pretty, levels = c("very low", "low", "medium", "high", "very high"))
absences.agg.to.plot.p$goout_pretty <- factor(absences.agg.to.plot.p$goout_pretty, levels = c("very low", "low", "medium", "high", "very high"))
```


```{r}
ggplot(absences.agg.to.plot.p, aes(x = freetime_pretty, y = goout_pretty)) +
  geom_tile(aes(fill = absences_p), color = 'white', show.legend = F) +
  theme_minimal() +
  geom_text(aes(label = absences_p), size = 5, fontface = 'bold', color = 'white') +
  labs(title ='How the amount of spare time and going out with friends \n influence the number of absences at Portugese classes?',
       x = "Free time after school",
       y = "Going out with friends") +
  theme(panel.grid = element_blank(),
        plot.title = element_text(hjust = 0.5, size = 15),
        axis.text = element_text(size = 10),
        axis.title = element_text(size = 15)) +
    scale_fill_gradient(low = color_palette$colors[3],
                        high = color_palette$colors[2])
```

## Insight

Most of the absences is related to the students that have medium amount of free time after school and are likely to go out with friends. It seems that hanging out with colleagues might influence more the number of absences than not having a lot of chores.

# Question 5

## Is parents' education and cohabitation status related to students' performance ?

```{r}
tmp1 <-   master_data %>%
  group_by(Medu,Fedu,Pstatus) %>%
  summarise(avg_grade_m = mean(G3_m) %>% round(2),
            avg_grade_p = mean(G3_p) %>% round(2)
            ) %>%
  mutate(grades = paste("M:",avg_grade_m,"\nP:",avg_grade_p))

master_data %>%
ggplot()+
  geom_bin2d(aes(as.factor(Medu),as.factor(Fedu), fill = ..count../nrow(master_data)))+
  labs(x = "mother's education",
       y = "father's education",
       fill = "fraction of students",
       title = "Average grades of students given their parents education")+
  scale_x_discrete(labels = c('none',
                              '4th grade',
                              '5th to 9th ngrade',
                              'secondary \neducation',
                              ' higher \neducation' ))+
    scale_y_discrete(labels = c('none',
                              '4th grade',
                              '5th to 9th grade',
                              'secondary \neducation',
                              ' higher education' ))+
  theme(axis.text.x = element_text(angle = 45, vjust = 0.6, hjust=0.5)) +
  geom_text(data = tmp1,aes(as.factor(Medu),as.factor(Fedu),label =grades),
            color = "white",
            size = 2,
            fontface = "bold")+
  scale_fill_gradient(low = color_palette$colors[3],
                      high = color_palette$colors[2])+
  facet_wrap(.~Pstatus,ncol=1 ,labeller = as_labeller(c("A" = "parents live apart","T" = "parents live together")))+
  theme(panel.background = element_blank(),
        panel.grid.major = element_line(colour = "grey"))
```

## Insight

We can infer that there is no clear relationship between parents education and
students' performance. Also parents' cohabitation status seems to be independent
from the performance. We can observe that most students have parents with same
level of education (red trend on diagonal).

# Question 6

## Is there any association between students' performance across periods and subjects?

```{r}
pass_fail_association <- master_data %>%
  mutate(current_failure_m = if_else(G3_m < 10, 1, 0),
         current_failure_p = if_else(G3_p < 10, 1, 0),
         failures_m_before = if_else(failures_m > 0, 1, 0),
         failures_p_before = if_else(failures_p > 0, 1, 0)) %>%
  group_by(failures_m_before, current_failure_m, failures_p_before, current_failure_p) %>% count()

pass_fail_association$group <- c("Rockstars",
                                 "Accidents \n will happen",
                                 "Accidents \n will happen",
                                 "Accidents \n will happen",
                                 "Accidents \n will happen",
                                 "Accidents \n will happen",
                                 "Accepting trade-off",
                                 "Rockstars",
                                 "Misery loves \n company",
                                 "Accepting trade-off",
                                 "Misery loves \n company",
                                 "Misery loves \n company",
                                 "Misery loves \n company")

pass_fail_association_aggregate <- pass_fail_association %>%
  group_by(group) %>%
  summarise(student_count = sum(n))
```

We created five groups that describe the consistency of the students' performance across periods.

- **The Rockstars** are depicted as the students that either are the gifted ones and pass both subjects consequently or after failing both previously now passed both of them.

- **Accidents will happen** describes a group of people that generally pass both Mathematics and Portugese subjects but have experiences at least one failure.

- With the **Accepting trade-off** cluster we describe the students that sacrifice one subject for the sake of the other.

- By **the mistery loves company** description we try to motivate the folks that have issues with learning at school and consequently experience problems at school.

```{r}
ggplot(pass_fail_association_aggregate,
       aes(area = student_count,
           label = group,
           fill = student_count)) +
  geom_treemap() +
  geom_treemap_text(fontface = "italic", colour = "white", place = "centre") +
  labs(title = "What groups students form across school years?",
       fill = "Students \n count") +
  theme(plot.title = element_text(hjust = 0.5, size = 15),
        legend.title = element_text(hjust = 0.5, size = 15),
        axis.text = element_text(size = 15)) +
  scale_fill_gradient(low = color_palette$colors[3],
                      high = color_palette$colors[2])
```

## Insight

There is a huge group of students that do not have problems with passing the subjects. Some people are aware of their
capabilities and accepted the trade-off. Luckily, the misery loves company is not very big. Considering the last group
it is bigger than it might be expected.

# Question 7

## What is the impact of students' lifestyle for their performance?

```{r}
all_numeric_var <- dictionary %>% filter(type == 'numeric') %>% pull(variable)

dims <- paste0(all_numeric_var[4:11], "_m")

dims <- append(dims, "G3_m")

final_dims <- c("Dalc_m", "Walc_m", "studytime_m", "absences_m", "G3_m")

master_data_to_radar_m <- master_data %>%
    mutate(across(all_of(final_dims), scales::rescale)) %>%
    group_by(higher_m) %>%
    summarise(across(all_of(final_dims), ~mean(.)))

names(master_data_to_radar_m) <- c("Wants to take \n higher education",
                                   "Workday alcohol \n consumption",
                                   "Weekend \n alcohol \n consumption",
                                   "Weekly \n study time",
                                   "Number of \n school absences",
                                   "Mathematics \n final grade")
```
```{r, dpi = 100}
ggradar(master_data_to_radar_m,
        plot.title = "What is the impact of students' lifestyle?",
        legend.title = "Wants to \n take higher \n education",
        axis.label.size = 3,
        legend.text.size = 10,
        gridline.mid.colour = color_palette$colors[4],
        group.colours = color_palette$colors[c(2,3)]) +
  theme(legend.title=element_text(size=10), plot.title = element_text(size = 15))
```

## Insight

Our assumptions are mostly correct and are confirmed with the graph. Students that want to continue their education
at universities have better grades in Mathematics and their lifestyle is better. Interestingly the number of school
absences does not play a remarkable role.

# Question 8

## Are there gender differences in grades?

```{r}
# calculate percentages
helper_df <- master_data %>%
  summarise(mop = sum(G3_m>G3_p),
            pom = sum(G3_m<G3_p),
            fmop = sum((G3_m>G3_p) & sex == 'F'),
            fpom = sum((G3_m<G3_p) & sex == 'F')) %>%
  mutate(fmop_perc = fmop/mop, mmop_perc = 1-fmop_perc,
         fpom_perc = fpom/pom, mpom_perc = 1-fpom_perc,
         mop_perc = mop/nrow(master_data),
         pom_perc = pom/nrow(master_data))

helper_df_2 <- helper_df %>%
  pivot_longer(c(mmop_perc,fmop_perc,mpom_perc,fpom_perc)) %>%
  select(c(name,value)) %>%
  .[1:2,]

helper_df_3 <- helper_df %>%
  pivot_longer(c(mmop_perc,fmop_perc,mpom_perc,fpom_perc)) %>%
  select(c(name,value)) %>%
  .[3:4,]

 seg_end_lst <- get_diag_segment_ends(side_length = 22)
 seg_end <- seg_end_lst$end
 fem_perc_diag <- seg_end_lst$fem_perc
                                                  
plt <-   
  ggplot()+
  geom_segment(aes(x=0,y=0,xend=seg_end,yend=seg_end), size = 6, color ="#F05454", alpha = 0.5 )+
  geom_segment(aes(x=seg_end,y=seg_end,xend=20,yend=20), size = 6, color ="#30475E",alpha = 0.5 )+
  geom_point(data = master_data,aes(x=G3_m,y=G3_p,color = sex),
             size = 3,
             position=position_dodge(width=0.5),
             alpha = 1)+
  annotate("text",label = paste("Better portuguese \nthan maths:\n",
                                round(helper_df$pom_perc*100),
                                "% of all"),
            size = 4 ,
           fontface = "italic",
           x =4.5, y =17)+
  annotate("text",
           label = paste("Better maths \nthan portuguese:\n",
                        round(helper_df$mop_perc*100),
                        "% of all"),
           size = 4 ,
           fontface = "italic",
           x =17, y =3) +
  annotate("text",
           label = paste0(round(fem_perc_diag*100),"%"),
           size = 4,
           color ="White",
           x =1, y =1,
           angle = 45)+
    annotate("text",
           label = paste0(round((1-fem_perc_diag)*100),"%"),
           size = 4,
           color ="White",
           x =19, y =19,
           angle = 45)+
  guides(fill="none")+
  scale_color_manual(values = c("#F05454","#30475E"),labels = c("female","male"))+
  coord_fixed()+
  labs(x="Maths grade", y="Portuguese grade", title = "Relationship between grades from two subjects by gender")+
  theme(panel.grid = element_blank(),
        legend.box.background = element_blank(),
        legend.key = element_rect(colour = "transparent",fill = alpha("red", 0)),
        plot.margin = margin(0,0,0,-20)
        )

  plt_b <- ggplot(data = helper_df_2)+
  geom_colh(aes(y = 0,x = value,fill = name),
            position = "stackv",
            width = 0.5)+
  scale_fill_manual(values = c("#F05454", "#30475E"),
                    labels = c("female","male")) + 
  theme_void()+
  theme(legend.position = "none")+
  geom_text(aes(x = (1-value)+0.1, y = 0,label = paste0(round(value,2)*100,"%")),
            color = "white")
  
  plt_l <- ggplot(data = helper_df_3)+
  geom_col(aes(x = 0,y = value,fill = name),
           position = "stack",
            width = 0.5)+
  scale_fill_manual(values = c("#F05454", "#30475E"),
                    labels = c("female","male")) + 
  theme_void()+
  theme(legend.position = "none")+
  geom_text(aes(x =0 , y = value - 0.1,label = paste0(round(value,2)*100,"%")),
            color = "white")

  lay <- rbind(c(1,2,2,2),
               c(NA,NA,NA,NA),
               c(NA,NA,3,NA))
    
  plt_fin <- arrangeGrob(plt_l,plt,plt_b, 
              layout_matrix = lay,
              widths = c(3,1,20,5),
              heights = c(25,1,2))

  grid.arrange(plt_fin)
 
```

## Insight (1)

We can clearly see that in general students perform better in Portuguese class
than in math class. In addition, among those who are better in maths there are
more males but in among those who are better in Portuguese - female are dominating.



```{r}
math_pyr <- master_data %>% 
  group_by(age,sex) %>% 
  summarise(grade_m = mean(G3_m), grade_p = mean(G3_p), n = n()) %>% 
  filter(age<=19) %>% 
  mutate(sex = ifelse(sex == "F","Female","Male")) %>%
  pyramid_chart( x = age, y = grade_m, group = sex,
               xlab = "Maths average grade",
               bar_colors = c("#F05454","#30475E"))

port_pyr <- master_data %>% 
  group_by(age,sex) %>% 
  summarise(grade_m = mean(G3_m), grade_p = mean(G3_p), n = n()) %>%
  filter(age<=19) %>%
  mutate(sex = ifelse(sex == "F","Female","Male")) %>% 

pyramid_chart( x = age, y = grade_p, group = sex,
               xlab = "Portuguese average grade",
               bar_colors = c("#F05454","#30475E"))

comb <- ggarrange(math_pyr,port_pyr)

annotate_figure(comb,top = text_grob("Grade average by age and gender",
                                       size = 20))
```

## Insight (2)

The distribution of grades among genders when controlling for age reveals that
the difference in performance in math classes is most striking among the youngest students.

# References