---
title: "AVR"
output:
  html_document:
      df_print: paged
      toc: yes
      toc_float: yes
      collapsed: yes
      smooth_scroll: no
bibliography: bibliography.bib
nocite: '@*'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(tidyverse, readxl, ggpubr, riverplot, gridExtra, grid, tidyverse, treemapify, ggradar, ggstance)
# options(device = "windows")
```

# Introduction

Data was gathered using questionnaires and mark reports. Mark reports were provided
by schools and consist of grades and absences. Grades are evaluated in three periods
and marked in a 20-point grading scale where 0 is the lowest possible mark and
20 the highest. The former was used to gather the data for the several demographic,
social/emotional and school related variables. The final questionnaire containing
37 questions in a single A4 sheet was answered in class by 788 students. Later
111 answers were discarded due to lack of identification details. At the end the
data was merged into two samples that related to Mathematics (395) and
Portuguese (649) classes.

The merged dataset consist of 13 variables that uniquely identifies
each row and 20 that describe their performance at one of the subjects. 
All the features can be split into categorical and numerical ones, but we 
are rather interested in analyzing the variables rather than the right survey distribution.

```{r}
index <- c("school","sex","age","address","famsize","Pstatus","Medu","Fedu","Mjob","Fjob","reason","nursery","internet")
index
```



```{r}
source("read_data.R")
source("plot_functions.R")
master_data <- read_data()
dictionary <- read_excel("dictionary.xlsx")
```

Let's plot the dependency between grades in Math and Portugese subjects.

```{r}
ggplot(master_data, aes(G1_m, G1_p)) +
  geom_point(aes(size = freetime_m)) +
  geom_smooth() +
  labs(x = "Grade in Mathematics",
       y = "Grade in Portugese",
       title = "Relationship between results in the first period")
```
Smaller points in the plot show that students with very little freetime haven't managed to keep excellent results in both subjects and they need to prioritize. 

# Question 1

What is the distribution of the health reported during both classes.

When analyzing the dataset it confused us that there are two separate columns for health.
We were convinced that this should be gathered only once during the research. The decision was made to check how attending the particular subject influences the responses.

```{r}

health_eq_acc_points <- master_data %>%
        mutate(health_eq = if_else(health_m == health_p, 1, 0)) %>%
        group_by(health_m, health_p) %>%
        summarise(cnt = n())


```


```{r}
ggplot(master_data %>%
               inner_join(health_eq_acc_points),
       aes(x = health_m, y = health_p, size = cnt)) +
        scale_size(name   = "",
                   breaks = c(144, 61, 1),
                   labels = c('144', '[42, 80]', '[1, 2]')) +
        geom_point() +
        labs(y = 'Health reported in the math classes',
             x = 'Health reported in the portugese classes',
             title = 'Relationship between reported health at Math and Portguese classess')
```


Most students reported the same health condition in Math and Portuguese classes.

# Question 2
What kind of support students with each subject? 

```{r barplots}

sum(master_data$paid_m != master_data$paid_p)
# subject support from family

master_data %>% select(famsup_m,famsup_p) %>% mutate(n=1) %>% pivot_longer(!n) %>% 
  ggplot(aes(x = value, fill = name)) +
  geom_bar(stat = "count", position = "dodge") +
  scale_fill_discrete(name = "Subject", labels = c("Maths", "Portuguese"))+
  ggtitle("Does student obtain family support?")+
  xlab(NULL)

# paid support
master_data %>% select(paid_m,paid_p) %>% mutate(n=1) %>% pivot_longer(!n) %>% 
  ggplot(aes(x = value, fill = name)) +
  geom_bar(stat = "count", position = "dodge")+
  scale_fill_discrete(name = "Subject", labels = c("Maths", "Portuguese"))+
  ggtitle("Does student obtain paid support?")+
  xlab(NULL)

# school support
master_data %>% select(schoolsup_m,schoolsup_p) %>% mutate(n=1) %>% pivot_longer(!n) %>% 
  ggplot(aes(x = value, fill = name)) +
  geom_bar(stat = "count", position = "dodge")+
  scale_fill_discrete(name = "Subject", labels = c("Maths", "Portuguese"))+
  ggtitle("Does student obtain school support?")+
  xlab(NULL)

```


```{r}
# better varsion of barplot to answer the quesition MATHS

# create additional variables

master_data %>% group_by(famsup_m,schoolsup_m,paid_m) %>% 
  summarise(n = n(), grade = mean(G3_m)) %>% 
  arrange(-grade) %>% pivot_longer(c(famsup_m,schoolsup_m,paid_m)) %>% 
  mutate(bg_color = ifelse(value == "yes", "green","red")) ->tmp_1

supp_m <- master_data %>% group_by(famsup_m,schoolsup_m,paid_m) %>% 
  summarise(n = n(), grade = mean(G3_m), .groups = "rowwise") %>% 
  mutate(cat_name = paste("f",famsup_m,"s",schoolsup_m,"p",paid_m,sep = '_')) %>% 
  mutate(id = cur_group_id())

tmp_1 <- supp_m  %>% pivot_longer(c(famsup_m,schoolsup_m,paid_m))
## tiles
# calculate positions of tiles
max_n <- 1#max(supp_m$n)
tile_half_height <- (max_n*1.01)/6
tile_y_middles <- c(tile_half_height,3*tile_half_height,5*tile_half_height)

tiles_df <- data.frame(
  x = rep(seq(1,8,1), 3),
  y = rep(tile_y_middles, each = 8)
)%>% arrange(x) %>% mutate(z = tmp_1$value) 

#maths
supp_plot_m <- ggplot() +
  geom_tile(data = tiles_df, aes(x, y,fill = z), colour = "white", alpha = 0.3)+
  scale_fill_manual(breaks=c("no","yes"),values=c("#f21a02","#a3d487"), name = "Support provided?")+
  scale_x_discrete()+
  geom_col(data = supp_m, aes(x=cat_name,y=n/sum(n)), alpha = 0.8, width = 0.8, fill = "steelblue")+
  scale_y_continuous(breaks = seq(0,1, length.out = 5) 
                    ,sec.axis = sec_axis( ~.,
                                          breaks =c(tile_half_height,3*tile_half_height,5*tile_half_height),
                                          labels = c("Family support","School support", "Paid support")))+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        panel.background = element_blank(),
        axis.line.y = element_line())+
  ylab("Fraction of students")+
  ggtitle("Maths")

supp_plot_m 

```

```{r}
# better varsion of barplot to answer the quesition PORT

# create additional variables

master_data %>% group_by(famsup_p,schoolsup_p,paid_p) %>% 
  summarise(n = n(), grade = mean(G3_p)) %>% 
  arrange(-grade) %>% pivot_longer(c(famsup_p,schoolsup_p,paid_p)) %>% 
  mutate(bg_color = ifelse(value == "yes", "green","red")) ->tmp_1

supp_p <- master_data %>% group_by(famsup_p,schoolsup_p,paid_p) %>% 
  summarise(n = n(), grade = mean(G3_p), .groups = "rowwise") %>% 
  mutate(cat_name = paste("f",famsup_p,"s",schoolsup_p,"p",paid_p,sep = '_')) %>% 
  mutate(id = cur_group_id())

tmp_1 <- supp_p  %>% pivot_longer(c(famsup_p,schoolsup_p,paid_p))
## tiles
# calculate positions of tiles
max_n <- max(supp_p$n)
tile_half_height <- (max_n+2)/6
tile_y_middles <- c(tile_half_height,3*tile_half_height,5*tile_half_height)

tiles_df <- data.frame(
  x = rep(seq(1,8,1), 3),
  y = rep(tile_y_middles, each = 8)
)%>% arrange(x) %>% mutate(z = tmp_1$value) 


supp_plot_p <- ggplot() +
  geom_tile(data = tiles_df, aes(x, y,fill = z), colour = "white", alpha = 0.3)+
  scale_fill_manual(breaks=c("no","yes"),values=c("#f21a02","#a3d487"), name = "Support provided?")+
  scale_x_discrete()+
  geom_col(data = supp_p, aes(x=cat_name,y=n), alpha = 0.8, width = 0.8, fill = "steelblue")+
  scale_y_continuous(breaks = seq(10,max(supp_p$n), length.out = 5) %>% round() 
                    ,sec.axis = sec_axis( ~.,
                                          breaks =c(tile_half_height,3*tile_half_height,5*tile_half_height),
                                          labels = c("Family support","School support", "Paid support")))+
  theme(axis.title.x=element_blank(),
        axis.text.x=element_blank(),
        axis.ticks.x=element_blank(),
        panel.background = element_blank(),
        axis.line.y = element_line())+
  ylab("Number of students")+
  ggtitle("Portuguese")


```
```{r}

supp_plot_m <- make_barplot_with_bg_tiles(master_data,famsup_m,schoolsup_m,paid_m,G3_m,
                                          title = '')
supp_plot_p <- make_barplot_with_bg_tiles(master_data,famsup_p,schoolsup_p,paid_p,G3_p)
supp_plot_m
supp_plot_p
ggarrange(supp_plot_p,supp_plot_m,common.legend = T , legend = "bottom")
```

# Question 3
Do students improve their grades during the year?

```{r}
# attempt 1
# data
tmp1 <- master_data %>% 
  select(matches("G._m")) %>% 
  drop_na()%>% 
  mutate(across(everything(),
                ~cut(.x,breaks = c(-1,7,15,21),
                    labels = c("low","medium","high")))) %>% 
  group_by(G1_m,G2_m) %>%
  summarise(n = n()) %>% 
  mutate(G1_m = paste0("First ",G1_m),
         G2_m = paste0("Second ",G2_m)) %>% 
  arrange(-n)


# final plot
nodes <- data.frame(ID = unique(c(tmp1$G1_m,tmp1$G2_m)),
                    x = c(rep(1,3),rep(2,3)))
edges <- data.frame(ID = 1:nrow(tmp1),
                    N1 = tmp1$G1_m,
                    N2 = tmp1$G2_m,
                    Value = tmp1$n)
cols <- c(low="#00990066",
medium="#00009966",
high ="#99000066")
style <- sapply(nodes$ID, function(id)
list(col=cols[ gsub("(First|Second) ", "", id) ]), simplify=FALSE)


r <- makeRiver(nodes,edges,styles =style) 
d <- list(srt=0, textcex=1.5) # default style
plot(r, plot_area=1, nodewidth=10, default_style=d)

```


# Question 4
What is the distribution of grades across subjects and grades in the current period?

```{r}
grade_per_subject_long <-
  master_data %>%
  mutate(row_num = row_number()) %>%
  select(row_num, starts_with("G", ignore.case = FALSE)) %>%
  pivot_longer(-row_num) %>%
  separate(name, c("grade", "subject"))
```

```{r}
subject_pretty_names <- tibble(short_name = c('m', 'p'),
                               pretty_name = c('Mathematics','Portugese'))
```


```{r}
ggplot(data = grade_per_subject_long %>%
  inner_join(subject_pretty_names,
             by = c("subject" = "short_name")) %>% separate(grade, 1, into = c("to_remove", "period")),
       aes(y = value, x = factor(pretty_name), fill = factor(period), colour=factor(period))) +
  geom_boxplot(alpha = 0.6, show.legend = F) +
  geom_point(position = position_jitterdodge()) +
  ggtitle("Grades from Mathematics and Portugese") +
  labs(x = "Subject", y = "Grade", colour = "Period") + guides(fill=FALSE)
```



Grades for Portugese were consistently higher than in Mathematics.
We do not observe neither a positive nor negative trend relating
to the subsequent periods results'.

# Question 5
How the amount of spare time and the way of spending it influence
the number of absences.


```{r}
grade_per_subject_long <-
  master_data %>%
  mutate(row_num = row_number()) %>%
  select(row_num, starts_with("G", ignore.case = FALSE)) %>%
  pivot_longer(-row_num) %>%
  separate(name, c("grade", "subject"))

absences.agg <- master_data %>%
  group_by(freetime_m, goout_m) %>%
  summarise(across(absences_m, ~n()))

low.high.map <- tibble(ugly_level = c(1, 2, 3, 4, 5),
                         pretty_level = c('very low',
                                               'low',
                                               'medium',
                                               'high',
                                               'very high'))

absences.agg.to.plot.m <- absences.agg %>%
  ungroup() %>%
  inner_join(low.high.map,
             by = c("freetime_m" = "ugly_level")) %>%
  rename( freetime_pretty = pretty_level) %>%
  inner_join(low.high.map,
             by = c("goout_m" = "ugly_level")) %>%
  rename( goout_pretty = pretty_level) %>% select(freetime_pretty, goout_pretty, absences_m)

absences.m.plot <-
  ggplot(absences.agg.to.plot.m, aes(x = freetime_pretty, y = goout_pretty)) + # nowa baza danych
  geom_tile(aes(fill = absences_m), color = 'white', show.legend = F) +
  theme_minimal() +
  geom_text(aes(label = absences_m), size = 5, fontface = 'bold', color = 'white') +
    theme(panel.grid = element_blank(),
          axis.title.x = element_blank(),
          axis.title.y = element_blank()) + ggtitle('Mathematics')
```
```{r}
absences.agg <- master_data %>%
  group_by(freetime_p, goout_p) %>%
  summarise(across(absences_p, ~n()))

low.high.map <- tibble(ugly_level = c(1, 2, 3, 4, 5),
                       pretty_level = c('very low',
                                        'low',
                                        'medium',
                                        'high',
                                        'very high'))
```


```{r}
absences.agg.to.plot.p <- absences.agg %>%
  ungroup() %>%
  inner_join(low.high.map,
             by = c("freetime_p" = "ugly_level")) %>%
  rename( freetime_pretty = pretty_level) %>%
  inner_join(low.high.map,
             by = c("goout_p" = "ugly_level")) %>%
  rename( goout_pretty = pretty_level) %>% select(freetime_pretty, goout_pretty, absences_p)
```


```{r}
absences.p.plot <-
ggplot(absences.agg.to.plot.p, aes(x = freetime_pretty, y = goout_pretty)) + # nowa baza danych
  geom_tile(aes(fill = absences_p), color = 'white', show.legend = F) +
  theme_minimal() +
  geom_text(aes(label = absences_p), size = 5, fontface = 'bold', color = 'white') +
  theme(panel.grid = element_blank(),
axis.title.x = element_blank(),
        axis.title.y = element_blank()) + ggtitle('Portugese')
```

```{r}
grid.arrange(arrangeGrob(absences.m.plot + theme(legend.position = "none"),
                         absences.p.plot + theme(legend.position = "none"),
                         ncol = 2),
             nrow = 2,
             heights = c(10, 1), # relative height of two rows
             top = textGrob("Consideration of the spare time might be an indicator for the number of absences.", vjust = 1, gp = gpar(fontface = "bold", cex = 1.5)),
             left = textGrob("Free time after school", rot = 90, vjust = 1),
             bottom = textGrob("Going out with friends", rot = 0, vjust = -2)
)

```

# Question 6
Does relationship status influence a number of absences?

```{r}
tmp1 <-   master_data %>% 
  group_by(Medu,Fedu,Pstatus) %>% 
  summarise(avg_grade_m = mean(G3_m) %>% round(2),
            avg_grade_p = mean(G3_p) %>% round(2) 
            ) %>% 
  mutate(grades = paste("Math:",avg_grade_m,"\nPort.:",avg_grade_p)) 

master_data %>% 
ggplot()+
  geom_bin2d(aes(as.factor(Medu),as.factor(Fedu), fill = ..count../nrow(master_data)))+
  labs(x = "mother's education",
       y = "father's education",
       fill = "% of students",
       title = "Average grades of students given their parents education")+
  scale_x_discrete(labels = c('none',
                              '4th grade',
                              '5th to 9th grade',
                              'secondary education',
                              ' higher education' ))+
    scale_y_discrete(labels = c('none',
                              '4th grade',
                              '5th to 9th grade',
                              'secondary education',
                              ' higher education' ))+ 
  theme(axis.text.x = element_text(angle = 45, vjust = 0.6, hjust=0.5)) + 
  geom_text(data = tmp1,aes(as.factor(Medu),as.factor(Fedu),label =grades),
            color = "white")+
  facet_wrap(.~Pstatus, labeller = as_labeller(c("A" = "parents live apart","T" = "parents live together")))
  
  master_data %>% 
  ggplot(aes(x=famrel_m,y=G3_m))+
    geom_point()+
    geom_smooth()
  
   master_data %>% 
  ggplot(aes(x=famrel_p,y=G3_p))+
    geom_point()+
    geom_smooth()



```
# Question 7
```{r}
pass_fail_association <- master_data %>%
  mutate(current_failure_m = if_else(G3_m >= 10, 1, 0),
         current_failure_p = if_else(G3_p >= 10, 1, 0),
         failures_m_general = if_else(failures_m > 0, 1, 0),
         failures_p_general = if_else(failures_p > 0, 1, 0)) %>%
  group_by(failures_m_general, current_failure_m, failures_p_general, current_failure_p) %>%
  count()
pass_fail_association$group <- c("Rockstars",
                                 "Accidents will happen",
                                 "Accidents will happen",
                                 "Achilles' heel",
                                 "Achilles' heel",
                                 "Accidents will happen",
                                 "Accepting trade-off",
                                 "Rockstars",
                                 "Misery loves company",
                                 "Achilles' heel",
                                 "Achilles' heel",
                                 "Misery loves company",
                                 "Misery loves company")
pass_fail_association_aggregate <- pass_fail_association %>%
  group_by(group) %>%
  summarise(student_count = sum(n))
```

We created five groups that describe the consistency of the students' performance across periods.

- Accidents will happen describe a group of people that generally pass both Mathematics and Portugese subjects but have experiences at least one failure.

- With the Achilles' heel we associate guys that consequently have problem with passing either Mathematics or Portugese.

- The Rockstars have been depicted as the students that either are the gifted ones and pass both subjects consequently or after failing both previously now passed them.

- With the mistery loves company description we try to motivate the folks that have issues with learning at school and consequently experience problems at school.

- Accepting trade-off relate to the students that sacrifice one subject for the sake of the other.

```{r}
ggplot(pass_fail_association_aggregate,
       aes(area = student_count,
           label = group,
       fill = student_count)) +
  geom_treemap() +
  geom_treemap_text(fontface = "italic", colour = "white", place = "centre",
                    grow = F) +
  labs(title = "Consistency of pass/failure across years") +
  scale_fill_gradient(guide = guide_colorbar(barheight = 20))
```

# Question 8
```{r}
all_numeric_var <- dictionary %>% filter(type == 'numeric') %>% pull(variable)
dims <- paste0(all_numeric_var[4:11], "_m")
dims <- append(dims, "G3_m")
final_dims <- c("Dalc_m", "Walc_m", "studytime_m", "absences_m", "G3_m")
```



```{r}
master_data_to_radar <- master_data %>%
    mutate(across(final_dims, scales::rescale)) %>%
    group_by(higher_m) %>%
    summarise(across(final_dims, ~mean(.)))
```
```{r}
names(master_data_to_radar) <- c("wants to take higher education", "workday alcohol consumption", "weekend alcohol consumption", "weekly study time", "number of school absences", "final grade")
```
```{r}
ggradar(master_data_to_radar)
```

# Question 9
Are there gender differences in grades?

```{r}
# calculate percentages
helper_df <- master_data %>%
  # mutate(fake_id = 1) %>% 
  # group_by(fake_id) %>% 
  summarise(mop = sum(G3_m>G3_p),
            pom = sum(G3_m<G3_p),
            fmop = sum((G3_m>G3_p) & sex == 'F'),
            fpom = sum((G3_m<G3_p) & sex == 'F')) %>% 
  mutate(fmop_perc = fmop/mop, mmop_perc = 1-fmop_perc,
         fpom_perc = fpom/pom, mpom_perc = 1-fpom_perc,
         mop_perc = mop/(mop+pom),
         pom_perc = 1- mop_perc)

helper_df_2 <- helper_df %>% 
  pivot_longer(c(mmop_perc,fmop_perc,mpom_perc,fpom_perc)) %>%
  select(c(name,value)) %>% 
  .[1:2,]

helper_df_3 <- helper_df %>% 
  pivot_longer(c(mmop_perc,fmop_perc,mpom_perc,fpom_perc)) %>%
  select(c(name,value)) %>% 
  .[3:4,]
                                                       
master_data %>% 
  #pivot_longer(c(G3_m,G3_p)) %>% 
  ggplot()+
  geom_colh(data = helper_df_2,aes(y = -1,x = 20*value,fill = name),width = 1,
             position = "stackv")+
  geom_col(data = helper_df_3,aes(x = -1,y = 20*value,fill = name),width = 1,
             position = "stack")+
  geom_point(aes(x=G3_m,y=G3_p,color = sex),
             size = 3,
             position=position_dodge(width=0.3))+
  geom_abline(slope = 1,intercept = 0)+
  annotate("text",label = paste("Better portuguese \nthan maths:\n",
                                round(helper_df$pom_perc*100),
                                "% of all"),
           x =5.5, y =17)+
  annotate("text",
           label = paste("Better maths \nthan portuguese:\n",
                        round(helper_df$mop_perc*100),
                        "% of all"),
           x =17, y =3) +
  scale_fill_manual(values = c("#F8766D", "#F8766D","#00BFC4", "#00BFC4"),
                    )+
  guides(fill="none")+
  scale_color_manual(values = c("#F8766D", "#00BFC4"),labels = c("female","male"))+
  coord_fixed()
  
  

   
    ggplot(data = helper_df_2)+
    geom_col(data = helper_df_2,aes(x = 100,y = value,fill = name),width = 0.4)
```




# References